Cognate beginnings to bilingual lexical acquisition

### 2023-06-28 manuscript/manuscript.qmd
Update PDF format

---
title: "Cognate beginnings to bilingual lexical acquisition"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias Fargas 25-27
        city: Barcelona
        region: Spain
        postal-code: "08005"
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        address: Alameda da Universidade
        city: Lisboa
        region: Portugal
        postal-code: "1600-214"
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        address: Iván Pavlov, 6
        city: Madrid
        region: Spain
        postal-code: "28049"
  - name: Núria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations: 
      - ref: cbc
date: "May 10th, 2023"
code-repo: "Access the code, data, and analysis at <https://github.com/gongcastro/trajectories>"
keywords:
  - lexical acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source
lang: en

format:
  pdf:
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header:
    - templates/include-in-header.tex
    header-includes: 
    - \usepackage{tipa}
    - \usepackage{nowidow}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DodgerBlue
    urlcolor: DodgerBlue
    citecolor: DodgerBlue
    link-citations: true
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true
  docx:
    knitr:
      opts_chunk:
        dev: "png"
        dev-args:
          - "cairo-png"
        dpi: 300
        out.width: "80%"
author-note:
  disclosures:
    conflict-of-interest: |
        The authors declare no conflicts of interest with regard to the funding source of this study.
    financial-support: |
        This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019).
    gratitude: |
        We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

@@

clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

@@

set.seed(888)

options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient,
        knitr.graphics.error = FALSE) 
```

The foundations of word learning are in place early in age. At six months of age, infants start directing their gaze to some objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and, shortly after, caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon show some interesting patterns.

@@

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated. The probability of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive additional activation through parallel activation more often than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language, roughly following a one-to-one mapping^[Although for some of the included words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the Catalan and Spanish lists]. (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*) (see @tbl-items for a summary of the questionnaire items).

```{r tbl-items}
#| label: tbl-items
#| tbl-cap: "Summary of the items included in the final analyses."
#| eval: true
#| echo: false
#| warning: false
items |> 
    arrange(language, list) |> 
    distinct(te, .keep_all = TRUE) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    group_by(semantic_category) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .groups = "drop") |> 
    mutate(total = rowSums(across(a:d)),
           items = map(items, function(x) head(x, 2))) |>  
    arrange(desc(total)) |> 
    relocate(a:d, total, items) |> 
    gt(rowname_col = "semantic_category") |> 
    fmt_percent(matches("_prop"), decimals = 1) |> 
    cols_label(a = "List A",
               b = "List B",
               c = "List C",
               d = "List D",
               total = "Total",
               items = "Examples") |> 
    tab_style(cell_text(size = "small"),
              list(cells_body(),
                   cells_stub(),
                   cells_column_labels(),
                   cells_column_spanners(),
                   cells_row_groups())) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_options(data_row.padding = px(2),
                row_group.padding = px(1.2)) |> 
    tab_style(cell_text(size = "x-small", align = "left"),
              cells_body(items))

@@

We manually generated a broad phonological transcription of every word included in the vocabulary checklists in X-SAMPA format [@wells1995computercoding]. Catalan word-forms were transcribed to Central Catalan phonology, and Spanish word-forms were transcribed to Castilian Spanish phonology. For each translation equivalent, we calculated the phonological similarity between the Catalan and the Spanish word-form using the Levenshtein distance, which measures the edit distance between each pair of phoneme strings [@levenshtein1966binary]. The output indicates the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other. We then divided this value by the length of the longest phoneme string and subtracted the outcome from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different, and one indicates that the two word-forms are identical [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological for similar approaches].

@@

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4-month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
ftn <- "For participants exposed to Catalan and Spanish exclusively this proportion is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    sub_missing(everything(), everything(), missing_text = "--") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:7) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan))) |>
    tab_footnote(ftn, location = cells_column_labels(doe_catalan))
```

@@

We collected `r format(nrow(n_items), big.mark = ",")` items responses. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to words `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. (1) *Age*, calculated as the number of months elapsed between participants' birth date and questionnaire completion). (2) *Length*, the number of phonemes in the X-SAMPA phonological transcription of the word-form. (3) *Exposure*, a language exposure-weighted lexical frequency. To account for the fact that bilinguals' exposure to a given word-form is a function of the word-forms lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We then mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We then multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure score to the Catalan word *cavall* [*horse*]---with a lexical frequency of `r round(items$freq[items$item=="cat_cavall"], 2)`---would be `r round(items$freq[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq[items$item=="cat_cavall"]*0.20, 2)`. (4) *Cognateness*, defined as the phonological similarity between translation equivalents, and calculated as detailed in the *Questionnaire* section.  These predictors were included as main effects in the model, together with their two-way and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data (see @eq-linear) [@barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}), \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2)
\end{aligned}
$$ {#eq-linear}

>>>>

---
title: "Cognate beginnings to bilingual lexical acquisition"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        city: Barcelona, 08005
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    email: avila.varela.daniela@gmail.com
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        city: Lisboa, 1600-214
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    email: jignaciocastillejo@gmail.com
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        city: Madrid, 28049
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    email: nuria.sebastian@upf.edu
    affiliations: 
      - ref: cbc
keywords:
  - cognate
  - word acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source
lang: en

format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: true
    doublespacing: true
    runninghead: "Cognate beginnings to bilingual lexical acquistion"
    include-in-header: 
      text: |
        \usepackage{tipa}
        \usepackage{booktabs}
abstract: |
    Bilingual infants show equivalent developmental trajectories of lexical acquisition compared to their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalise on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

@@

@@

set.seed(888)
```

# Introduction[^author-note]

[^author-note]: The authors declare no conflicts of interest with regard to the funding source of this study. This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

The foundations of word learning are in place early in age. At six months, infants start directing their gaze to some objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early], and shortly after caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon show some interesting patterns.

@@

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated^[We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not make strong assumptions about whether a learning instance is a discrete or a continuous unit of accumulation of information. We rather consider that a learning instance of a word is so to the degree of the strength of activation of its phonological representation. This activation may be the result of the infant being exposed to the word-form, or the result of activation spreading through phonological or semantic links across lexicl representation, as in the case of parallel activation.]. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive additional activation through parallel activation more often than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (OCDI) [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping^[Although for some of the included words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the Catalan and Spanish lists]  (see @tbl-items for a summary of the questionnaire items).

```{r tbl-items}
#| label: tbl-items
#| tbl-cap: "Summary of the items included in the final analyses."
#| eval: true
#| echo: false
#| warning: false
items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(total = rowSums(across(a:d)),
           items = map(items, function(x) tail(x, 2))) |>  
    arrange(desc(total)) |> 
    relocate(a:d, total, items) |> 
    mutate(group = "Semantic category") |> 
    select(-total) |> 
    gt(groupname_col = "group",
       rowname_col = "semantic_category") |> 
    fmt_percent(matches("_prop"), decimals = 1) |> 
    cols_label(a = "List A",
               b = "List B",
               c = "List C",
               d = "List D",
               items = "Examples") |> 
    summary_rows(groups = "Semantic category",
                 columns = a:d,
                 fns = list("Total" = ~sum(., na.rm = TRUE))
    ) |> 
    tab_style(cell_text(size = "small"),
              list(cells_body(),
                   cells_summary(),
                   cells_stub(),
                   cells_column_labels(),
                   cells_column_spanners(),
                   cells_stub_summary(),
                   cells_row_groups())) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_options(data_row.padding = px(2),
                row_group.padding = px(1.2)) |> 
    tab_style(cell_text(size = "x-small", align = "left"),
              cells_body(items))

@@

To compute predictors of interest, we manually generated a broad phonological transcription of every word included in the vocabulary checklists in X-SAMPA format [@wells1995computercoding]. Catalan word-forms were transcribed to Central Catalan phonology, and Spanish word-forms were transcribed to Castilian Spanish phonology.

@@

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4-month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
ftn <- "For participants exposed to Catalan and Spanish exclusively this proportion is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    sub_missing(everything(), everything(), missing_text = "-") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:7) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan))) |>
    tab_footnote(ftn, location = cells_column_labels(doe_catalan))
```

@@

We collected responses to `r format(nrow(n_items), big.mark = ",")` items. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to words `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest: *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (a language exposure-weighted lexical frequency), and *Cognateness* (defined as the phonological similarity between translation equivalentsFor each translation equivalent). We added these variables as main effects, together with the two-way and three-way interactions between *Age*, *Exposure*, and *Cognateness*.

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-forms lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We then mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We then multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure score to the Catalan word *cavall* [*horse*]---with a lexical frequency of `r round(items$freq[items$item=="cat_cavall"], 2)`---would be `r round(items$freq[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq[items$item=="cat_cavall"]*0.20, 2)`.

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levnehstein distance of each translation equivalent by the length of the longest word-form to correct for word length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /'taw.l5-'me.sa/, *table*), and one indicates that the two word-forms are identical (e.g., /'mar-'mar/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological for similar approaches].

Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [@barr2013random]. We specified a weakly informative prior around the parameters of the model. @eq-linear shows a detailed description of the model.

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}) \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\text{where:}\\
& u_{1-8_{i}}~\text{: participant-level adjustments} \\
& w_{1-3_{j}}~\text{: TE-level adjustments} \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2) \\
&\text{where }~\rho_{u_{0-8}, w_{0-3}}~\text{are the correlations between group-level adjustments}\\
\end{aligned}
$$ {#eq-linear}
